Virtualisierung eines Echtzeit-Betriebssystems zur Steuerung eines Roboters mit Schwerpunkt auf die Einhaltung der Echtzeit

### 2024-04-05 Masterarbeit_Robotik_Pamuk.tex
Write masterthesis

%- Literaturverzeichnis
%- Abbildungsverzeichnis
%- Tabellenverzeichnis
%- Abkürzungsverzeichnis (optional)
%- Glossar (optional)
%- Anhang (optional)
\chapter{Introduction}
In today's industrial production and automation, robot systems are well established and of crucial importance. Robots must react to their environment and perform time-critical tasks within strict time constraints. Delays or errors can have catastrophic consequences in some cases. Traditional operating systems, such as Windows or Linux, are often not suitable for these types of real-time requirements as they cannot guarantee deterministic execution times. Therefore, real-time operating systems are required that are specifically designed to react to events within fixed time limits and prioritise the execution of high-priority processes.

@@

\clearpage
\section{Application Context}
This master's thesis was written at SIGMATEK GmbH \& Co KG \cite{pixelartSIGMATEKKompletteAutomatisierungssysteme}. 
%This section provides necessary information about the starting point and boundary conditions of the thesis, as the work should be reproducible with the means used. 
SIGMATEK uses its own customized Linux distribution to be run on their self-manufactured CPUs, namely Salamander 4. This operating system employs hard real-time with latency requirements between 20 and 50 \textmu s. The goal is to virtualize Salamander 4 and approach the performance of bare metal CPUs. Salamander 4 is virtualized through a third party service, QEMU. The details of this operating system are explained in section \ref{sec:salamander4}.

\clearpage
\section{State of the art}
\clearpage

\section{Problem and task definition}

\clearpage
\section{Objective}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\chapter{Methodology}

@@

\clearpage
\chapter{Salamander Operating System}\label{sec:salamander4}

Salamander 4 is the proprietary operating system of SIGMATEK. It is based on Linux version 5.15.94 and integrates Xenomai 3.2, a real-time development environment \cite{XenomaiXenomai}. Salamander 4 is a 64-bit system, which refers to the x86\_64 architecture. The real-time behaviour is achieved through the use of Symmetric Multi-Processing (SMP) and Preemptive Scheduling (PREEMPT). In addition, it uses IRQPIPE to process interrupts in a way that meets the real-time requirements of the system. The output of the command \texttt{uname -a} can be observed in code \ref{lst:uname_a}.

\vspace{1em}
\begin{minipage}{0.95\columnwidth}
	\begin{lstlisting}[name={System information},label={lst:uname_a}]
root@sigmatek-core2:~# uname -a
Linux sigmatek-core2 5.15.94 #1 SMP PREEMPT IRQPIPE Tue Feb 14 18:18:05 UTC 2023 x86_64 GNU/Linux
\end{lstlisting}
\end{minipage}

@@

\section{Task priorities}
Es gibt grundsätzlich 4 Gruppen

\begin{table}[ht]
	\centering
	\caption{Overview of the priority groups and their relationships}
	\label{tab:priorities}
	\begin{tabular}{|c|c|}
		\hline
		\textbf{Prioritätsgruppe}    & \textbf{Bereich} \\ \hline
		Xenomai Priorität            & 0 bis 99         \\ \hline
		Linux RT Priorität           & 1 bis 99         \\ \hline
		Linux (Nice Level) Priorität & -20 bis 19       \\ \hline
		RTK Priorität                & 0 bis 14         \\ \hline
	\end{tabular}
\end{table}

\section{Memory Management}
Es gibt verschiedene Speicherbereiche

@@

\chapter{Comparison of Real-Time Latency}
In the initial phase, a comparative latency analysis was conducted between the hardware version and the virtualized version of Salamander 4. For this purpose, the latency tool of the Xenomai test suite was used. The latency was measured under two conditions, idle and CPU-stressed. The goal was to optimize the latency of the virtualisation of Salamander 4 OS to closely match that of the bare metal version.

After analyzing the inital latency of both versions, Trace-cmd and Kernelshark were used to further inspect the reasons that caused this divergence.

@@

Figure \ref{fig:max_latency_hardware} shows latency of hardware Salamander4.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\columnwidth]{img/max_latency_hardware.png}
	\caption[Latency hardware]{Latency hardware}
	\label{fig:max_latency_hardware}
\end{figure}

\section{Salamander 4 Virtualisation}
In addition to providing Salamander 4 on its own hardware, SIGMATEK has also developed a virtualised version of this operating system. It was developped using Yocto, an open source project that allows customised Linux distributions to be created for embedded systems \cite{WelcomeYoctoProject}. The virtualisation runs in a QEMU environment, which is an open source tool for hardware virtualisation \cite{QEMU}. With the help of the script depicted in code \ref{lst:qemu_def}, Salamander 4 is started together with the necessary hardware components in the QEMU environment. This makes it possible to run Salamander 4 on a variety of host systems, regardless of the specific hardware of the host. Upon generating the necessary files, Yocto generates a QEMU folder with the following components shown in code \ref{lst:ls}.

\vspace{1em}
\begin{minipage}{\linewidth}
	\begin{lstlisting}[name={Contents of QEMU folder for Salamander 4},label={lst:ls}]
    sigma_ibo@localhost:~/Desktop/salamander-image$ ls -1
    bzImage
    drive-c
    ovmf.code.qcow2
    qemu_def_bkp.sh
    qemu_def.sh
    salamander-image-sigmatek-core2.ext4
    stek-drive-c-image-sigmatek-core2.tar.gz
    vmlinux
    \end{lstlisting}
\end{minipage}

\vspace{1em}
\begin{minipage}{\linewidth}
	\begin{lstlisting}[name={QEMU script for starting Salamander 4 virtualisation},label={lst:qemu_def}]
    #!/bin/sh

@@

    exec  qemu-system-x86_64 -M pc,accel=kvm -kernel ./bzImage \
    -m 2048 -drive file=salamander-image-sigmatek-core2.ext4,format=raw,media=disk \
    -append "console=ttyS0 console=tty1 root=/dev/sda rw panic=1 sigmatek_lrt.QEMU=1 ip=dhcp rootfstype=ext4 schedstats=enable" \
    -net nic,model=e1000,netdev=e1000 -netdev bridge,id=e1000,br=nm-bridge \
    -fsdev local,security_model=none,id=fsdev0,path=drive-c -device virtio-9p-pci,id=fs0,fsdev=fsdev0,mount_tag=/mnt/drive-C \
    -device vhost-vsock-pci,guest-cid=3,id=vsock0 \
    -drive if=pflash,format=qcow2,file=ovmf.code.qcow2 \
    -no-reboot -nographic
\end{lstlisting}
\end{minipage}

Here is a description of the used components:
\begin{itemize}
	\item \textbf{bzImage}: Compressed Linux kernel image, loaded by QEMU at system start.
	\item \textbf{ovmf.code.qcow2}: Firmware file for QEMU, enables UEFI boot process.
	\item \textbf{qemu\_def.sh}: Shell script, starts QEMU with correct parameters to boot Salamamder 4 OS.
	\item \textbf{stek-drive-c-image-sigmatek-core2.tar.gz}: Archive containing files for C drive, unpacked and copied to drive-c/ directory by qemu\_def.sh script.
	\item \textbf{drive-c}: Directory, serves as C drive for QEMU system, created and filled by qemu\_def.sh script..
	\item \textbf{salamander-image-sigmatek-core2.ext4}: Root file system for Salamander 4 OS, used as hard drive for QEMU system.
	\item \textbf{vmlinux}: Uncompressed Linux kernel image, typically used for debugging, contains debugging symbols not present in bzImage.
\end{itemize}

@@

\clearpage
\noindent When the script is started from the host, the QEMU process can be scheduled to run on any available core, as it is noted bound to a specific CPU core. This means that the QEMU process may frequently switch between different cores, leading to an increase in latency. As the goal was to reduce latency in the guest, the first step was to isolate a CPU of the host and dedicate it solely to the QEMU process, so that it cannot be used for other tasks on user level. However, the isolcpus function only isolates at the user level and does not affect kernel tasks. Consequently, these kernel tasks and interrupts can still utilize the CPU.

@@

Figure \ref{fig:gnuplot_max_latency_with_taskset}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/with_taskset/gnuplot_max_latency_with_taskset.png}
	\caption[gnuplot latency with taskset]{gnuplot latency with taskset}
	\label{fig:gnuplot_max_latency_with_taskset}
\end{figure}

\chapter{Latency Reduction}
\section{CPU isolation}
\clearpage
\section{User and Kerneltasks}
\clearpage
\section{IRQ}
\clearpage

\chapter{KVM exit reasons}
\section{APIC\_WRITE}

@@

\subsection{APIC virtualization}

@@

\clearpage
\section{HLT}
\subsection{Description}
\subsection{Occurence Reduction}
\clearpage
\section{EPT\_MISCONFIG}
\subsection{Description}
\subsection{Occurence Reduction}
\clearpage
\section{PREEMPTION\_TIMER}
\subsection{Description}
\subsection{Occurence Reduction}
\clearpage
\section{EXTERNAL\_INTERRUPT}
\subsection{Description}
\subsection{Occurence Reduction}
\clearpage
\section{IO\_INSTRUCTION}
\subsection{Description}
\subsection{Occurence Reduction}
\clearpage
\section{EOI\_INDUCED}
\subsection{Description}
\subsection{Occurence Reduction}
\clearpage
\section{EPT\_VIOLATION}
\subsection{Description}
\subsection{Occurence Reduction}
\clearpage
\section{PAUSE\_INSTRUCTION}
\subsection{Description}
\subsection{Occurence Reduction}
\clearpage
\section{CPUID}
\subsection{Description}
\subsection{Occurence Reduction}
\clearpage
\section{MSR\_READ}
\subsection{Description}
\subsection{Occurence Reduction}
\clearpage

\chapter{Results}

@@

\clearpage
\chapter{Discussion}
\clearpage
\chapter{Summary and Outlook}

@@

\phantomsection
\addcontentsline{toc}{chapter}{\listacroname}
\chapter*{\listacroname}
\begin{acronym}[XXXXX]
	\acro{CPU}[CPU]{Central Processing Unit}
	\acro{QEMU}[QEMU]{Quick Emulator}
\end{acronym}

%
% Hier beginnt der Anhang.
%
\clearpage
\appendix
\chapter{Anhang A}
\clearpage
\chapter{Anhang B}
\end{document}

>>>>

%- Literaturverzeichnis
%- Abbildungsverzeichnis
%- Tabellenverzeichnis
%- Abkürzungsverzeichnis (optional)
%- Glossar (optional)
%- Anhang (optional)
\chapter{Introduction}\label{cha:introduction}
In today's industrial production and automation, robot systems are well established and of crucial importance. Robots must react to their environment and perform time-critical tasks within strict time constraints. Delays or errors can have catastrophic consequences in some cases. Traditional operating systems, such as Windows or Linux, are often not suitable for these types of real-time requirements as they cannot guarantee deterministic execution times. Therefore, real-time operating systems are required that are specifically designed to react to events within fixed time limits and prioritise the execution of high-priority processes.

@@

\clearpage
\section{Application Context}
This master's thesis was written at SIGMATEK GmbH \& Co KG \cite{pixelartSIGMATEKKompletteAutomatisierungssysteme}.
%This section provides necessary information about the starting point and boundary conditions of the thesis, as the work should be reproducible with the means used. 
SIGMATEK uses its own customized Linux distribution to be run on their self-manufactured CPUs, namely Salamander 4. This operating system employs hard real-time with latency requirements between 20 and 50 \textmu s. The goal is to virtualize Salamander 4 and approach the performance of bare metal CPUs. Salamander 4 is virtualized through a third party service, QEMU. The details of this operating system are explained in chapter \ref{cha:salamander4}.

\clearpage
\section{State of the art}
\clearpage
\section{Problem and task definition}
\clearpage
\section{Objective}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage

\chapter{Methodology}\label{cha:methodology}

@@

\chapter{Salamander 4}\label{cha:salamander4}

Salamander 4 is the proprietary operating system of SIGMATEK. It is based on Linux version 5.15.94 and integrates Xenomai 3.2, a real-time development environment \cite{XenomaiXenomai}. Salamander 4 is a 64-bit system, which refers to the x86\_64 architecture. The real-time behaviour is achieved through the use of Symmetric Multi-Processing (SMP) and Preemptive Scheduling (PREEMPT). In addition, it uses IRQPIPE to process interrupts in a way that meets the real-time requirements of the system. The output of the command \texttt{uname -a} can be observed in code \ref{output:uname_a}.

\vspace{1em}
\begin{minipage}{0.95\columnwidth}
	\begin{lstlisting}[name={System information},label={output:uname_a}]
root@sigmatek-core2:~# uname -a
Linux sigmatek-core2 5.15.94 #1 SMP PREEMPT IRQPIPE Tue Feb 14 18:18:05 UTC 2023 x86_64 GNU/Linux
\end{lstlisting}
\end{minipage}

@@

\clearpage
\section{Salamander 4 Description}

\subsection{Task priorities}
Es gibt grundsätzlich 4 Gruppen

\begin{table}[ht]
	\centering
	\caption{Overview of the priority groups and their relationships}
	\label{tab:priorities}
	\begin{tabular}{|c|c|}
		\hline
		\textbf{Prioritätsgruppe}    & \textbf{Bereich} \\ \hline
		Xenomai Priorität            & 0 bis 99         \\ \hline
		Linux RT Priorität           & 1 bis 99         \\ \hline
		Linux (Nice Level) Priorität & -20 bis 19       \\ \hline
		RTK Priorität                & 0 bis 14         \\ \hline
	\end{tabular}
\end{table}
\clearpage

\subsection{Memory Management}
Es gibt verschiedene Speicherbereiche

@@

@@

Figure \ref{fig:max_latency_hardware} shows latency of hardware Salamander4.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\columnwidth]{img/max_latency_hardware.png}
	\caption[Latency hardware]{Latency hardware}
	\label{fig:max_latency_hardware}
\end{figure}
\clearpage

\section{Salamander 4 Virtualisation}
In addition to providing Salamander 4 on its own hardware, SIGMATEK has also developed a virtualised version of this operating system. It was developped using Yocto, an open source project that allows customised Linux distributions to be created for embedded systems \cite{WelcomeYoctoProject}. The virtualisation runs in a QEMU environment, which is an open source tool for hardware virtualisation \cite{QEMU}. With the help of the script depicted in code \ref{script:qemu_def}, Salamander 4 is started together with the necessary hardware components in the QEMU environment. This makes it possible to run Salamander 4 on a variety of host systems, regardless of the specific hardware of the host. Upon generating the necessary files, Yocto generates a QEMU folder with the following components shown in code \ref{lst:ls}.

\vspace{1em}
\begin{minipage}{\linewidth}
	\begin{lstlisting}[name={Contents of QEMU folder for Salamander 4},label={lst:ls}]
    sigma_ibo@localhost:~/Desktop/salamander-image$ ls -1
    bzImage
    drive-c
    ovmf.code.qcow2
    qemu_def.sh
    salamander-image-sigmatek-core2.ext4
    stek-drive-c-image-sigmatek-core2.tar.gz
    vmlinux
    \end{lstlisting}
\end{minipage}

\vspace{1em}
\begin{minipage}{\linewidth}
	\begin{lstlisting}[name={QEMU script for starting Salamander 4 virtualisation},label={script:qemu_def}]
    #!/bin/sh

@@

    exec  qemu-system-x86_64 -M pc,accel=kvm -kernel ./bzImage \
    -m 2048 -drive file=salamander-image-sigmatek-core2.ext4,format=raw,media=disk \
    -append "console=ttyS0 console=tty1 root=/dev/sda rw panic=1 sigmatek_lrt.QEMU=1 ip=dhcp rootfstype=ext4 schedstats=enable" \
    -net nic,model=e1000,netdev=e1000 -netdev bridge,id=e1000,br=nm-bridge \
    -fsdev local,security_model=none,id=fsdev0,path=drive-c -device virtio-9p-pci,id=fs0,fsdev=fsdev0,mount_tag=/mnt/drive-C \
    -device vhost-vsock-pci,guest-cid=3,id=vsock0 \
    -drive if=pflash,format=qcow2,file=ovmf.code.qcow2 \
    -no-reboot -nographic
\end{lstlisting}
\end{minipage}
\clearpage
Here is a description of the used components:
\begin{itemize}
	\item \textbf{bzImage}: Compressed Linux kernel image, loaded by QEMU at system start.
	\item \textbf{ovmf.code.qcow2}: Firmware file for QEMU, enables UEFI boot process.
	\item \textbf{qemu\_def.sh}: Shell script, starts QEMU with correct parameters to boot Salamamder 4 OS.
	\item \textbf{stek-drive-c-image-sigmatek-core2.tar.gz}: Archive containing files for C drive, unpacked and copied to drive-c/ directory by qemu\_def.sh script.
	\item \textbf{drive-c}: Directory serving as C drive for QEMU system, created and filled by qemu\_def.sh script.
	\item \textbf{salamander-image-sigmatek-core2.ext4}: Root file system for Salamander 4 OS, used as hard drive for QEMU system.
	\item \textbf{vmlinux}: Uncompressed Linux kernel image, typically used for debugging, contains debugging symbols not present in bzImage.
\end{itemize}

@@

\noindent When the script is started from the host, the QEMU process can be scheduled to run on any available core, as it is noted bound to a specific CPU core. This means that the QEMU process may frequently switch between different cores, leading to an increase in latency. As the goal was to reduce latency in the guest, the first step was to isolate a CPU of the host and dedicate it solely to the QEMU process, so that it cannot be used for other tasks on user level. However, the isolcpus function only isolates at the user level and does not affect kernel tasks. Consequently, these kernel tasks and interrupts can still utilize the CPU.

@@

Figure \ref{fig:gnuplot_max_latency_with_taskset}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/with_taskset/gnuplot_max_latency_with_taskset.png}
	\caption[gnuplot latency with taskset]{gnuplot latency with taskset}
	\label{fig:gnuplot_max_latency_with_taskset}
\end{figure}
\clearpage

\section{Latency Comparison}
In the initial phase, a comparative latency analysis was conducted between the hardware version and the virtualized version of Salamander 4. For this purpose, the latency tool of the Xenomai test suite was used. The latency was measured under two conditions, idle and CPU-stressed. The goal was to optimize the latency of the virtualisation of Salamander 4 OS to closely match that of the bare metal version.

After analyzing the inital latency of both versions, Trace-cmd and Kernelshark were used to further inspect the reasons that caused this divergence.
\clearpage

\chapter{Real-Time Latency Reduction}\label{cha:latency_reduction}
\section{CPU isolation}
Isolating CPUs inolves removing all user-space threads and unbound kernel threads since bound kernel threads are tied to specific CPUs and hence cannot be moved. Also, modifying the \texttt{proc/irq/IRQ\_NUMBER/smp\_affinity} property of each Interrupt IRQ\_NUMBER in the system is part of this process.

Output \ref{output:pse} shows

\vspace{1em}
\begin{minipage}{0.95\columnwidth}
	\begin{lstlisting}[name={User and Kernel Tasks},label={output:pse},escapeinside={(*@}{@*)}]
		sigma_ibo@sigma-ibo:~$ cat /sys/devices/system/cpu/isolated
		19
		sigma_ibo@sigma-ibo:~$ ps -e -o pid,psr,comm | awk '$2 == 19'
				92  19 cpuhp/19
				93  19 idle_inject/19
 	    		94  19 migration/19
 	    		95  19 ksoftirqd/19
 	    		97  19 kworker/19:0H-events_highpri
 	 		 1025  19 irq/205-iwlwifi:queue_7
 	 		17448  19 kworker/19:1H-kblockd
 	 		17499  19 kworker/19:2-events
 	 		18761  19 kworker/19:3-events
			21401  19 qemu-system-x86
\end{lstlisting}
\end{minipage}

\clearpage
\section{Kernel tuning}
\clearpage
\section{IRQ affinity}

The IRQ affinity can be configured so that interrupts are handled on CPUs other than CPU 19. This helps reduce the number of kernel threads that are executed on CPU 19.

The script \ref{script:irq_affinity} checks the smp\_affinity for each IRQ on a specific CPU and prints the IRQ numbers that are allowed to run on that CPU, sorted in ascending order.

\vspace{1em}
\begin{minipage}{0.95\columnwidth}
	\begin{lstlisting}[name={Read IRQ affinity per CPU on system},label={script:irq_affinity}]
		#!/bin/bash
		# Get the CPU number from the command-line argument
		CPU=$1
		echo -n "CPU $CPU: "
		for IRQ in /proc/irq/*; do
			if [ -f "$IRQ/smp_affinity" ]; then
				# Read the current smp_affinity
				AFFINITY=$(cat "$IRQ/smp_affinity")
				# Check if the bit for the current CPU is set
				if (( (0x$AFFINITY & (1 << CPU)) != 0 )); then
					# Print the IRQ number without a newline, followed by a space
					echo -n "${IRQ#/proc/irq/} "
				fi
			fi
		done | sort -n
		# Print a final newline
		echo
\end{lstlisting}
\end{minipage}

Output \ref{output:irq_affinity} shows the output of the script above for CPU 19.

\vspace{1em}
\begin{minipage}{0.95\columnwidth}
\begin{lstlisting}[name={Output},label={output:irq_affinity}, breaklines=true]
	sigma_ibo@sigma-ibo:~$ ./check_smp_affinity.sh 19
	CPU 19 IRQ affinity:
	0
	2
	3
	4
	5
	6
	7
	10
	11
	13
	15
	130
	172
	173
	189
	192
\end{lstlisting}
\end{minipage}

\clearpage
\section{KVM exit reasons}\label{sec:kvm_exit_reasons}
\subsection{APIC\_WRITE}

@@

\subsubsection{APIC virtualization}

@@

\clearpage
\subsection{HLT}
\subsubsection{Description}
\subsubsection{Occurence Reduction}
\clearpage
\subsection{EPT\_MISCONFIG}
\subsubsection{Description}
\subsubsection{Occurence Reduction}
\clearpage
\subsection{PREEMPTION\_TIMER}
\subsubsection{Description}
\subsubsection{Occurence Reduction}
\clearpage
\subsection{EXTERNAL\_INTERRUPT}
\subsubsection{Description}
\subsubsection{Occurence Reduction}
\clearpage
\subsection{IO\_INSTRUCTION}
\subsubsection{Description}
\subsubsection{Occurence Reduction}
\clearpage
\subsection{EOI\_INDUCED}
\subsubsection{Description}
\subsubsection{Occurence Reduction}
\clearpage
\subsection{EPT\_VIOLATION}
\subsubsection{Description}
\subsubsection{Occurence Reduction}
\clearpage
\subsection{PAUSE\_INSTRUCTION}
\subsubsection{Description}
\subsubsection{Occurence Reduction}
\clearpage
\subsection{CPUID}
\subsubsection{Description}
\subsubsection{Occurence Reduction}
\clearpage
\subsection{MSR\_READ}
\subsubsection{Description}
\subsubsection{Occurence Reduction}
\clearpage

\chapter{Results}\label{cha:results}

@@

\clearpage
\chapter{Discussion}\label{cha:discussion}
\clearpage
\chapter{Summary and Outlook}\label{cha:summary_and_outlook}

@@

\phantomsection
\addcontentsline{toc}{chapter}{\listacroname}
\chapter*{\listacroname}
\begin{acronym}[XXXXX]
	\acro{CPU}[CPU]{Central Processing Unit}
	\acro{QEMU}[QEMU]{Quick Emulator}
	\acro{IRQ}[Interrupt Request]{Interrupt Request}
\end{acronym}

%
% Hier beginnt der Anhang.
%
\clearpage
\appendix

\chapter{Anhang A}
\clearpage

\chapter{Anhang B}

\end{document}
